---
title: "R Notebook"
output: 
    html_notebook: default
    html_document: default
    
---

```{r}
#original template code for markdown adjusted for the use with Rnotebook
#---
#title: "Reproducible Research: Peer Assessment 1"
#output: 
#  html_document:
#    keep_md: true
#---

```

This report is written in a R notebook using Rstudio. 

To reproduces this notebook the following packages are called (you may need Rtools 
to correctly compile the required packages if not allready installed):

```{r results='hide'}
library(dplyr)
library(tidyr)
library(lubridate)
library(lattice)
library(kableExtra)
```


The file is downloaded from the [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
given in the assignment.

```{r results='hide'}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file (fileUrl, destfile="activity.zip", method="curl")
```

After downloading, the file is needs to be unzipped before loading into R.  

```{r}
unzip('activity.zip', exdir=".")
```

## Loading and preprocessing the data

The data is loaded into a dataframe using read.csv.

```{r}
dfa <- read.csv("activity.csv", header=TRUE, stringsAsFactors = FALSE)
str(dfa)  

summary(dfa)
```

The data is ready to proces and doens't need further processing. 


## What is the mean total number of steps taken per day?
The total number of steps taken is calculated as the sum of 
the 'steps' columns. 

```{r}
sumsteps <- summarise(dfa, totalsteps=sum(steps, na.rm=TRUE))
sumsteps$totalsteps
```



To review the steps by day, the data is aggregated by day. A histogram is created 
to show the frequencies of steps taken daily.  

```{r}
stepsbyday <- dfa %>%
        group_by(date) %>%
        summarise(stepsbd=sum(steps, na.rm = TRUE))

hist(stepsbyday$stepsbd, xlab="Steps per day", main="Histogram of daily steps")
```

The mean  and median  number
of steps taken daily are determined from the dataframe 'stepsbyday' created in 
the previous step. The values are presented in the table below. 


```{r}
summarydf <- summarise(stepsbyday,
        totalsteps=sum(stepsbd),
        meansteps=mean(stepsbd),
        mediansteps=median(stepsbd))

summarydf
```

## What is the average daily activity pattern?

To review the activity pattern by the 5 minute interval averaged by day, 
first the raw data is aggregated to daily averages by the 5 minute intervals. 

```{r}
stepsby5 <- dfa %>%
        drop_na() %>%
        group_by(interval) %>%
        summarise(meansteps=mean(steps))
head(stepsby5)
```

The average daily pattern is shown in a time series plot below. 

```{r}
plot(stepsby5$interval, stepsby5$meansteps, type="l", xlab="Time (by 5 min intervals)",
     ylab="average number of steps", main="Average number of steps by time stamp")
```

The maximum amount of steps taken daily on average is found at 
time stamp shown in the value below. 

```{r}
maxsteps5 <- max(stepsby5$meansteps)
maxinterval <- as.numeric(stepsby5[stepsby5$meansteps==maxsteps5,"interval"])
maxinterval
```




##Impute missing values

The summary of the data at the beginning of the document showed that only the 
variable steps contained NAs. The number of NAs in the dataset is shown below

```{r}
summarise(dfa, naCount=sum(is.na(steps)))
```

A simple approach to impute the missing values would be to assign the average value
of steps for the interval timestamp.

```{r}
dfaImputed <- dfa %>%
        left_join(y=stepsby5, by=c("interval"="interval")) %>%
        mutate(steps = as.double(steps))%>%
        mutate(steps = if_else(!is.na(steps), steps, meansteps))

summary(dfaImputed)
```

With the new dataset a new histogram can be drawn showing the frequincies of steps 
taken per day.

```{r}
stepsbydayImputed <- dfaImputed %>%
        group_by(date) %>%
        summarise(stepsbd=sum(steps, na.rm = TRUE))

hist(stepsbydayImputed$stepsbd, xlab="Steps per day", main="Histogram of daily steps
     with missing data imputed")
```

The imputation clearly affects the pattern of daily steps as does it affects the 
mean and median ammount of steps taken per day, as shown in the table below.


```{r}
summarydfImputed <- summarise(stepsbydayImputed,
        totalstepsImp=sum(stepsbd),
        meanstepsImp=mean(stepsbd),
        medianstepsImp=median(stepsbd))

summarydfImputed

```

Comparing the raw and the new imputed dataset, the mean increases with 
`r text_spec(round(summarydfImputed$meanstepsImp-summarydf$meansteps), color="red")` and 
the median increases with 
`r text_spec(round(summarydfImputed$medianstepsImp-summarydf$mediansteps), color="red")`.

## Are there differences in activity patterns between weekdays and weekends?
To answer the question,  for each date it has to be determined if it is a 
weekday or a weekend day. The function weekdays() is used to create a new variable
called dayoftheweek, which is used to create the variable daytype.

```{r}
curr_locale <- Sys.getlocale("LC_TIME")
dfaImputedWD <- dfaImputed %>%
    mutate(dayoftheweek = wday(as.Date(date), label=TRUE, abbr=FALSE, locale='us'),
           daytype = as.factor(if_else(dayoftheweek %in% c('Monday', 'Tuesday',
                            'Wednesday', 'Thursday', 'Friday'), 'weekday',
                            'weekend'))) %>%
# To be able to produce the requested graph the data is again grouped and summmarized
    group_by(interval, daytype) %>% 
    summarize(steps=mean(steps))
```

The new dataset is used to plot the mean number of steps taken at each 
5 min. interval during weekdays and weekend days. 

```{r}
xyplot(steps ~ interval | daytype, data=dfaImputedWD, layout=c(1,2), type=c("g", "l"),
       main="Average number of steps by 5 minute intervals")
```




